---
title: "Final Project Log"
author: "niall martin"
date: "5/12/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#House Prices Data
library(sf)
library(dplyr)


## Subsetting for 2019
directory <- getwd()
London_B <- st_read("data_opens_3.geojson")
View(London_B)




## Renaming variable for merge
London_B$pcds <- London_B$pcd2

```


```{r}
# Conversion Codes

conv <- read.csv("/Users/niallmartin22/Documents/DataSets/Conversion Codes.csv",stringsAsFactors = FALSE)
head(conv)
library(dplyr)
conv2 <- select(conv, pcds, lsoa11cd, lsoa11nm)
conv2$LSOA.Code <- conv2$lsoa11cd

```

Crime Dataset
```{r}
crime <- read.csv("/Users/niallmartin22/Documents/DataSets/HistoricCrime.csv",stringsAsFactors = FALSE)

##subsetting crime to 2011 & subset columns
crime$total <- rowSums(crime[41:52])
crime <- crime %>% select(LSOA.Code,Borough,total)
```


## Merging Datasets
```{r}
total <- merge(crime,conv2,by="LSOA.Code")
total2 <- merge(total,London_B, by="pcds")
```

Cleaning the dataset and subsetting 

```{r}
total2 <- total2[, colSums(is.na(total2)) == 0]

total2 <- total2 %>% select(pcds,LSOA.Code,Borough,total,price,propertytype,oldnew,tfarea,numberrooms,Dist_OpenSpace,geometry)
```


Calculating total crime for individual crime &
House price average per Borough 

```{r}

library(dplyr)
## Average crime per borough 
new_data <- total2 %>% group_by(Borough) %>%
  mutate(avg_crime1=mean(total))

##Average crime per borough 
#new_data$avg_crime <- new_data %>% select(total)%>% group_by(Borough) %>%
 # mutate(mean(total))

new_data2 <- new_data %>% select(price,tfarea,numberrooms,Dist_OpenSpace,avg_crime1)

new_data2[,c("Borough")] <- list(NULL)

```

```{r}
##Checking the distribution of continuous variables
library(ggplot2)
##Price
p <- ggplot(new_data, aes(x=price, y=price)) + 
  geom_violin()+
  coord_flip()
p

#Log
p1 <- ggplot(new_data2, aes(x=log(price), log(price))) + 
  geom_violin()+
  coord_flip()
p1

#tfarea
p2 <- ggplot(new_data, aes(x=tfarea, y=tfarea)) + 
  geom_violin()+
  coord_flip()
p2

#Log
p3 <- ggplot(new_data2, aes(x=log(tfarea), y=log(tfarea))) + 
  geom_violin()+
  coord_flip()
p3

#Inputting Log Values
new_data2$price<-log(new_data2$price)
new_data2$price<-log(new_data2$tfarea)

```

Hedonic model 

```{r}
library(Hmisc)
library(corrplot)
library(caret)

set.seed(1234)
inTrain <- createDataPartition(new_data2$price, p = 0.5, list = FALSE)
trainData <- new_data2[inTrain, ]
testData <- new_data2[-inTrain, ]
dim(trainData)

trainData <- as.data.frame(lapply(trainData, as.numeric))
testData <- as.data.frame(lapply(testData, as.numeric))

trainMatrix <- rcorr(as.matrix(trainData))
print(trainMatrix)

flattenCorrMatrix <- function(cormat, pmat){
  ut <- upper.tri(cormat)
  data.frame(row = rownames(cormat)[row(cormat)[ut]],
             column = rownames(cormat)[col(cormat)[ut]],
             cor = (cormat)[ut],
             p = (pmat)[ut])}
flattentrainMatrix <- flattenCorrMatrix(trainMatrix$r, trainMatrix$P)
print(flattentrainMatrix)


```

There is a positive correlation (0.7) between price and floor area. 

There is a positive correlation (0.5) between price and number of rooms. 

There is a negative correlation (-0.03) between price and distance to open area. 

There is a positive correlation (0.2) between price and average crime. 

```{r}
corrplot(trainMatrix$r, type = "upper", order = "FPC", method = "color",
         p.mat = trainMatrix$P, sig.level = 0.01, insig = "pch",
         tl.cex = 0.8, tl.col = "black", tl.srt = 45)
```

```{r}
# Building our Multiple Linear Regression Model
modFit <- train(price ~ Dist_OpenSpace + tfarea +
                  avg_crime1 + numberrooms, method = "lm", data = trainData)
finMod <- modFit$finalModel
print(modFit)
```

```{r}
print(finMod)
```

Regression Diagnostic plots 

```{r}
par(mfrow = c(2,2))
plot(finMod)
```

```{r}
##Linearity
plot(finMod, 1, pch = 19, cex = 0.5)
```


Normality of the Data 

```{r}
# Plotting the Normal Q-Q Plot
plot(finMod, 2, pch = 19, cex = 0.5)
```

Homoscedasticity

```{r}
# Plotting the Scale-Location Plot
plot(finMod, 3, pch = 19, cex = 0.5)

```

Outliers, High Leverage Points and Influential Values
```{r}
# Plotting the Residuals vs Leverage Plot
par(mfrow = c(1,2))
plot(finMod, 4, pch = 19, cex = 0.5)
plot(finMod, 5, pch = 19, cex = 0.5)
```

Prediction with our Multiple Linear Regression Model
```{r}
# Prediction with our Multiple Linear Regression Model
Prediction <- predict(modFit, testData)
qplot(price, Prediction, colour = Dist_OpenSpace, data = testData)
```

```{r}
qplot(price, Prediction, colour = numberrooms, data = testData)
```

```{r}
qplot(price, Prediction, colour = avg_crime1, data = testData)

```

